---
title: "Linking different metrics to Elaboration and Innovation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 8
    fig_height: 8
---


## Elaboration and Innovation definitions

For any species with coordinates $\vec{species}$ in a $n$ dimensional space, we can estimate $VCV_{superorder}$ and $VCV_{order}$, respectively two nested variance-covariance matrices representing the any order and super order that contains $\vec{species}$.
We can then calculate the eigen vector and eigenvalues from these matrices as: 

$\textbf{v}_{superorder}$ and $\lambda_{superorder}$ defined as:
$$VCV_{superorder} \textbf{v}_{superorder} = \lambda_{superorder} \textbf{v}_{superorder}$$

and $\textbf{v}_{order}$ and $\lambda_{order}$ defined as:
$$VCV_{order} \textbf{v}_{order} = \lambda_{order} \textbf{v}_{order}$$

In practice, these are solved using a eigen decomposition in `R` `eigen`.

We can then define the followings:

$$\text{Elaboration}_{\text{species}} = \|\frac{\vec{species} \cdot \textbf{v}_{order}} {\sqrt{\textbf{v}_{order} \cdot \textbf{v}_{order}}}\|$$

$$\text{Innovation}_{\text{species}} = \| \vec{species} - \text{Elaboration}_{\text{species}} \|$$

$$\text{Elaboration}_{\text{clade}} = \| \frac{\textbf{v}_{order} \cdot \textbf{v}_{superorder}} {\sqrt{\textbf{v}_{superorder} \cdot \textbf{v}_{superorder}}}\|$$

$$\text{Innovation}_{\text{clade}} = \| \textbf{v}_{order} - \text{Elaboration}_{\text{clade}} \|$$

```{r}
library(dispRity)

## Calculating the Elaboration/Innovation for clades and species
data(charadriiformes)

## Creating a dispRity object with a covar component
my_covar <- MCMCglmm.subsets(
                 data       = charadriiformes$data,
                 posteriors = charadriiformes$posteriors,
                 tree       = charadriiformes$tree,
                 group      = MCMCglmm.levels(
                                 charadriiformes$posteriors)[1:4],
                 rename.groups = c("gulls", "plovers", "sandpipers", "phylo"))

## Clade level
clade   <- dispRity.covar.projections(my_covar, type = "groups", base = "phylo", n = 1000,  output = c("position", "distance"))
species   <- dispRity.covar.projections(my_covar, type = "elements", n = 100,  output = c("position", "distance"))
```

## Relative eigenvalues dispersion (Watanabe, 2022)

$$V_{rel} = \frac{\sum_{i}^{n}(\lambda_{i} - \sum{\frac{\lambda_{i}}{n}})^2}{n(n-1)\sum{\frac{\lambda_{i}}{n}}^2}$$
Which is the dispersion of eigenvalues (squared difference between each eigen value and the average eigen value).


```{r, eval = TRUE}
## Relative eigenvalue dispersion
rel.eigenvalue.dispersion <- function(matrix, get.var = FALSE) {
    if(get.var) {
        matrix <- var(matrix)
    }
    ## Getting the eigen values, n and mean eigen value
    lambdas <- eigen(matrix)$values
    n <- length(lambdas)
    lambda_mean <- mean(lambdas)
    ## Calculating the relative eigen dispersion
    return(sum((lambdas - lambda_mean)^2)/n*(n-1)*lambda_mean)
}

## Calculate the eigen dispersions
eigen_dispersion <- dispRity(my_covar, metric = as.covar(rel.eigenvalue.dispersion))

## Get the values
clade_disp <- get.disparity(eigen_dispersion, concatenate = FALSE)
clade_elab <- get.disparity(clade$position, concatenate = FALSE)
clade_inov <- get.disparity(clade$distance, concatenate = FALSE)

par(mfrow = c(1,2))
plot(NULL, xlim = range(unlist(clade_disp)), ylim = range(unlist(clade_elab)), xlab = "Eigen dispersion", ylab = "Elaboration")
points(clade_disp$gulls, clade_elab$gulls, pch = 19, col = "darkgreen")
points(clade_disp$sandpipers, clade_elab$sandpipers, pch = 19, col = "blue")
points(clade_disp$plovers, clade_elab$plovers, pch = 19, col = "orange")
legend("topright", pch = 19, col = c("darkgreen", "blue", "orange"), legend = c("gulls", "sandpipers", "plovers"))

plot(NULL, xlim = range(unlist(clade_disp)), ylim = range(unlist(clade_inov)), xlab = "Eigen dispersion", ylab = "Innovation")
points(clade_disp$gulls, clade_inov$gulls, pch = 19, col = "darkgreen")
points(clade_disp$sandpipers, clade_inov$sandpipers, pch = 19, col = "blue")
points(clade_disp$plovers, clade_inov$plovers, pch = 19, col = "orange")
```

## Conditional evolvability (Hansen & Houle, 2008)

If species can be interpreted as "the response to selection in $n$ traits", then:

$$\text{Evolvability} = \text{Elaboration}_{\text{species}}$$

$$\text{Respondability} = \sqrt{\text{Elaboration}_{\text{species}}^2 + \text{Innovation}_{\text{species}}^2}$$

Conditional evolvability = evolvability * autonomy
Autonomy = fraction of genetic variation that is independent of potentially constraining characters = inverse of G matrix.


```{r, eval = TRUE}
## Relative eigenvalue dispersion
respondability <- function(x, y) {
    sqrt(x^2 + y^2)
}

## calculating the respondability
species_elab <- get.disparity(species$position, concatenate = FALSE)
species_inov <- get.disparity(species$distance, concatenate = FALSE)
species_resp <- mapply(respondability, species_elab, species_inov, SIMPLIFY = FALSE)

par(mfrow = c(1,2))
plot(NULL, xlim = range(unlist(species_resp)), ylim = range(unlist(species_elab)), xlab = "Respondability", ylab = "Elaboration (evolvability)")
points(species_resp$plovers, species_elab$plovers, pch = 19, col = "orange", cex = 0.5)
points(species_resp$sandpipers, species_elab$sandpipers, pch = 19, col = "blue", cex = 0.5)
points(species_resp$gulls, species_elab$gulls, pch = 19, col = "darkgreen", cex = 0.5)
legend("topright", pch = 19, col = c("darkgreen", "blue", "orange"), legend = c("gulls", "sandpipers", "plovers"))

plot(NULL, xlim = range(unlist(species_resp)), ylim = range(unlist(species_inov)), xlab = "Respondability", ylab = "Innovation")
points(species_resp$plovers, species_inov$plovers, pch = 19, col = "orange", cex = 0.5)
points(species_resp$sandpipers, species_inov$sandpipers, pch = 19, col = "blue", cex = 0.5)
points(species_resp$gulls, species_inov$gulls, pch = 19, col = "darkgreen", cex = 0.5)
```





 - [ ] link orthogonality to random skewers
 - [x] link E/I to relative eigenvector dispersion (Pavlicev et al., 2009, Watanabe, 2022) 
 - [x] link E/I to conditional evolvability (Hansen & Houle, 2008) 
 - [ ] link E/I to integration (Hansen & Houle, 2008) 

