---
title: "Bird beak's elaboration and innovation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

This section contains the code for reproducing the analyses. To get the details, refer to [`dispRity` vignette](link_to_projection_vignette).
This script take around 6 minutes to compile.

```{r, eval = TRUE, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, print = FALSE}
library(dispRity)
source("../Functions/ploting.utilities.R")
```

# All birds shapespace analyses

```{r, eval = TRUE, echo = FALSE}
## Load the data
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
## Select the data
shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

## Load the covar
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]

## Get the levels
phylo_lvl <- which(MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order) == "animal")
super_lvl <- grep("level1_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
order_lvl <- grep("level2_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))

## Creating the subsets
superorder_shapespace <- MCMCglmm.subsets(n = 4000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_allbirds_lvl_superorder_order,
      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(super_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""

order_shapespace <- MCMCglmm.subsets(n = 4000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_allbirds_lvl_superorder_order,
      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(order_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)
```

First we can visualise the global bird shape space (or at least the 2 first of the 8 dimensions):

```{r, message = FALSE, warning = FALSE, fig.height = 8, fig.width = 8, echo = FALSE}
# colour.palette <- rainbow
colour.palette <- gg.color.hue

## Plotting just the space
covar.plot(superorder_shapespace,
           main = "Shapespace",
           cex = 0.5,
           legend = TRUE,
           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
           legend.args = list(x = "bottomright", cex = 2/3))
```

Figure: The global shape space


We can then check the distribution of the variance covariance matrices for each super order and each order.
The VCV matrices are represented as average ellipses or average axis.

```{r, echo = FALSE, fig.height = 14, fig.width = 14, eval = TRUE}
par(mfrow = c(2,2))
## Plotting mean axes
covar.plot(superorder_shapespace,
           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
           main = "Average ellipses by super order",
           ellipses = mean,
           legend = FALSE,
           points = FALSE,
           cex = 0.5, apply.to.VCV = TRUE)
## Plotting mean ellipses
covar.plot(superorder_shapespace,
           col = c(colour.palette(n.subsets(superorder_shapespace)-1), "grey"),
           main = "Average major axes by super order",
           major.axes = mean,
           legend = TRUE,
           points = FALSE,
           cex = 0.5, apply.to.VCV = TRUE)
covar.plot(order_shapespace,
           col = c(colour.palette(n.subsets(order_shapespace)-1), "grey"),
           main = "Average ellipses by order",
           ellipses = mean,
           legend = FALSE,
           points = FALSE,
           cex = 0.5, apply.to.VCV = TRUE)
## Plotting mean ellipses
covar.plot(order_shapespace,
           col = c(colour.palette(n.subsets(order_shapespace)-1), "grey"),
           main = "Average major axes by order",
           major.axes = mean,
           legend = TRUE,
           points = FALSE,
           cex = 0.5, apply.to.VCV = TRUE)
```

Main axes and ellipses visualisation for each superorders and orders.

### Are these 2D average ellipses and/or axes an accurate representation of what's going on? 
<details>
  <summary>[Click to expand] TL;DR: It depends...
</summary>
  
A couple of things have to be taken into account:

 * First, the averaging of the VCV (and it's ellipse and/or maon axis) can result into weird "middle" VCV that are not representing the individual VCV (e.g. if half VCV are negative, and half positive, the average will be probably 0; which does not exist).
 * Second, the ellipses in nD projected onto a 2D space will be deformed (unless the VCV is a real pancake on the first 2D, i.e. with no variance on the other dimensions).

Note that both problems are only visual though since: 1) we don't use averages anywhere (but the full distributions all the time) and 2) the linear algebra bit with the projections and stuff works in any number of dimensions and transforms the results into 1D results (i.e. the projection/rejection of groups/species on the major axes in 8D space).

Here's a visualisation of how the variance in VCVs affects the averaging artefact:

```{r, eval = TRUE, echo = FALSE}
visualisation.group(superorder_shapespace, "Mirandornithes")
```

```{r, echo = FALSE, eval = TRUE}
visualisation.group(superorder_shapespace, "Strisores")
```

Or even more obvious (with an direction that does not really exist (due to averaging)):

```{r, echo = FALSE, eval = TRUE}
visualisation.group(superorder_shapespace, "Paleognathae")
```

For the 2Dness, the effect is a bit less obvious (because the two first dimensions already have so much variance). Here is an example for the Columbimorphae. (to check different groups, you can easily modify the source code of the Rmd. TODO: this would be good in some shiny app innovation thingy).


```{r, echo = FALSE, eval = TRUE, fig.height = 6, fig.width = 12}
par(mfrow = c(1,2))
visualisation.group(superorder_shapespace, "Columbimorphae")
visualisation.group(superorder_shapespace, "Columbimorphae", dimensions = c(1,2))
```

Back to the main story.
</details>

## Calculate the elaborations/innovations

We then calculate the projections/rejections as explained in more details in the [03-elaboration_exploration_analyses.Rmd]() script.
Really briefly we calculate three types of projections/rejections abbreviated as follows below:

 * `group on phylogeny`: the projection of each group's VCV (major axis) on the phylogenetic's VCV (major axis)
 * `species on phylogeny`: the projection of each element within a group (species) onto the phylogenetic's VCV (major axis)
 * `species on group`: the projection of each element within a group (species) onto its own VCV (major axis)


<!-- TODO: add the pictures of the different tests here -->


```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}
## Run the analyses between groups
super_results_between <- dispRity.covar.projections(superorder_shapespace,
                                             type = "groups",
                                             base = "phylogeny",
                                             verbose = TRUE,
                                             output = c("position", "distance"), centre = TRUE, abs = TRUE)
## Run the analyses between groups
order_results_between <- dispRity.covar.projections(order_shapespace,
                                             type = "groups",
                                             base = "phylogeny",
                                             verbose = TRUE,
                                             output = c("position", "distance"), centre = TRUE, abs = TRUE)
## Run the analyses of groups on the phylogeny
super_results_phylo <- dispRity.covar.projections(superorder_shapespace,
                                            type = "elements",
                                            base = "phylogeny",
                                            inc.base = TRUE,
                                            verbose = TRUE,
                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)

## Run the analyses of groups on the phylogeny
order_results_phylo <- dispRity.covar.projections(order_shapespace,
                                            type = "elements",
                                            base = "phylogeny",
                                            inc.base = TRUE,
                                            verbose = TRUE,
                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
## Run the analyses groups on themselves
super_results_within <- dispRity.covar.projections(superorder_shapespace,
                                            type = "elements",
                                            verbose = TRUE,
                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
## Run the analyses groups on themselves
order_results_within <- dispRity.covar.projections(order_shapespace,
                                            type = "elements",
                                            verbose = TRUE,
                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
shapespace_allbirds_lvl_superorder_order_results_list <- list(
        "super_results_between" = super_results_between,
        "order_results_between" = order_results_between,
        "super_results_phylo"   = super_results_phylo,
        "order_results_phylo"   = order_results_phylo,
        "super_results_within"  = super_results_within,
        "order_results_within"  = order_results_within)
save(shapespace_allbirds_lvl_superorder_order_results_list, file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
```

```{r, echo = FALSE}
## Quick load
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
results <- shapespace_allbirds_lvl_superorder_order_results_list
rm(shapespace_allbirds_lvl_superorder_order_results_list)
```


## Saving plotting data (median values)


```{r, echo = FALSE, eval = FALSE}
## Getting the plotting data
ploting_data <- list("median_elaborations" = apply(get.disparity(get.subsets(results$super_results_phylo$position, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
                     "median_innovations"  = apply(get.disparity(get.subsets(results$super_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
                     "distances_to_centre" = get.disparity(dispRity(get.subsets(superorder_shapespace, subsets = "phylogeny"), metric = dispRity::centroids, centroid = 0))[[1]],
                     "consensus_tree"      = tree) 

## Get the data at the macro level (with no missing data)
get.macro.projection <- function(data, what) {
    if(what == "centroids") {
        ## Measuring the centroid distances
        data <- dispRity(data, metric = dispRity::centroids)
    } else {
        data <- data[[what]]
    }
    ## Getting all the species
    all_species <- apply(get.disparity(get.subsets(data, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median)
    names(all_species) <- rownames(get.subsets(data, subsets = "phylogeny")$matrix[[1]])
    ## Getting the species that have a projection on their own group
    macro_level <- apply(do.call(rbind, get.disparity(get.subsets(data, subsets = names(size.subsets(data))[-n.subsets(data)]), concatenate = FALSE)), 1, median)
    names(macro_level) <- unlist(lapply(lapply(get.subsets(data, subsets = names(size.subsets(data))[-n.subsets(data)])$subsets, `[[`, "elements"), function(X, matrix) rownames(matrix)[X], matrix = data$matrix[[1]]))
    ## Replacing the species that have their own group
    all_species[match(names(macro_level), names(all_species))] <- macro_level
    return(all_species)
}

## Getting the plotting data super_order_level
ploting_data_superorder <- list(
                     "median_elaborations" = get.macro.projection(results$super_results_within, what = "position"),
                     "median_innovations"  = get.macro.projection(results$super_results_within, what = "distance"),
                     "distances_to_centre" = get.macro.projection(superorder_shapespace, what = "centroids"),
                     "consensus_tree"      = tree) 

## Getting the plotting data super_order_level
ploting_data_order <- list(
                     "median_elaborations" = get.macro.projection(results$order_results_within, what = "position"),
                     "median_innovations"  = get.macro.projection(results$order_results_within, what = "distance"),
                     "distances_to_centre" = get.macro.projection(order_shapespace, what = "centroids"),
                     "consensus_tree"      = tree) 

## Save all that as a list
save(ploting_data, file = "../Data/Processed/ploting_data.rda")
save(ploting_data_superorder, file = "../Data/Processed/ploting_data_superorder.rda")
save(ploting_data_order, file = "../Data/Processed/ploting_data_order.rda")

## Group info
super_orders <- lapply(results$super_results_phylo$position$subsets, function(x, data) list(species = rownames(data)[c(x$elements)]), data = results$super_results_phylo$position$matrix[[1]])
orders <- lapply(results$order_results_phylo$position$subsets, function(x, data) list(species = rownames(data)[c(x$elements)]), data = results$order_results_phylo$position$matrix[[1]])
## Adding the colours
load("../Data/Processed/tip_colours_super_orders.rda")
load("../Data/Processed/tip_colours_orders.rda")
add.col <- function(one_order, one_col) {
    one_order$col <- one_col
    return(one_order)
}
super_orders <- mapply(add.col, one_order = super_orders[1:8][match(names(super_orders[1:8]), names(tip_colours_super_orders))], one_col = as.list(tip_colours_super_orders), SIMPLIFY = FALSE)
orders <- mapply(add.col, one_order = orders[1:27][match(names(orders[1:27]), names(tip_colours_orders))], one_col = as.list(tip_colours_orders), SIMPLIFY = FALSE)

## Big list:
clade_info <- list(super_orders = super_orders,
                   orders = orders)
## Save all that data
save(clade_info, file = "../Data/Processed/clade_info.rda")

## Get the names
names(clade_info$orders)

## Get the colours
unlist(lapply(clade_info$orders, `[[`, "col"))

## Get the species in one specific group
clade_info$orders$Cuculiformes$species

```

# Passeriformes shapespace analyses

```{r, eval = TRUE, echo = FALSE}
## Load the data
load(file = "../Data/Processed/shapespace_passeriformes_two_levels.rda")
## Select the data
shapespace <- shapespace_passeriformes_two_levels[[1]]$space
tree <- shapespace_passeriformes_two_levels[[1]]$consensus_tree
## Clean garbage
rm(shapespace_passeriformes_two_levels)

## Load the covar
load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
shapespace_passeriformes <- shapespace_passeriformes[[1]]

## Get the levels
phylo_lvl <- which(MCMCglmm.levels(shapespace_passeriformes) == "animal")
suborder_lvl <- grep("level1_", MCMCglmm.levels(shapespace_passeriformes))
family_lvl <- grep("level2_", MCMCglmm.levels(shapespace_passeriformes))

## Creating the subsets
suborder_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_passeriformes,
      group         = MCMCglmm.levels(shapespace_passeriformes)[c(suborder_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""

family_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_passeriformes,
      group         = MCMCglmm.levels(shapespace_passeriformes)[c(family_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_passeriformes)
```

## Calculate the elaborations/innovations

```{r, fig.height = 20, fig.width = 10, echo = FALSE, eval = FALSE, results = 'hide', message = FALSE, warning = FALSE}
## Run the analyses between groups
suborder_results_between <- dispRity.covar.projections(suborder_shapespace,
                                             type = "groups",
                                             base = "phylogeny",
                                             verbose = TRUE,
                                             output = c("position", "distance"), centre = TRUE, abs = TRUE)
## Run the analyses between groups
family_results_between <- dispRity.covar.projections(family_shapespace,
                                             type = "groups",
                                             base = "phylogeny",
                                             verbose = TRUE,
                                             output = c("position", "distance"), centre = TRUE, abs = TRUE)
## Run the analyses of groups on the phylogeny
suborder_results_phylo <- dispRity.covar.projections(suborder_shapespace,
                                            type = "elements",
                                            base = "phylogeny",
                                            inc.base = TRUE,
                                            verbose = TRUE,
                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)

## Run the analyses of groups on the phylogeny
family_results_phylo <- dispRity.covar.projections(family_shapespace,
                                            type = "elements",
                                            base = "phylogeny",
                                            inc.base = TRUE,
                                            verbose = TRUE,
                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
## Run the analyses groups on themselves
suborder_results_within <- dispRity.covar.projections(suborder_shapespace,
                                            type = "elements",
                                            verbose = TRUE,
                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
## Run the analyses groups on themselves
family_results_within <- dispRity.covar.projections(family_shapespace,
                                            type = "elements",
                                            verbose = TRUE,
                                            output = c("position", "distance"), centre = TRUE, abs = TRUE)
shapespace_passeriformes_suborder_family_results_list <- list(
        "suborder_results_between" = suborder_results_between,
        "family_results_between"   = family_results_between,
        "suborder_results_phylo"   = suborder_results_phylo,
        "family_results_phylo"     = family_results_phylo,
        "suborder_results_within"  = suborder_results_within,
        "family_results_within"    = family_results_within)
save(shapespace_passeriformes_suborder_family_results_list, file = "../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")
```

## Saving plotting data (median values)


```{r, echo = FALSE, eval = FALSE}
load("../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")
results <- shapespace_passeriformes_suborder_family_results_list
rm(shapespace_passeriformes_suborder_family_results_list)
## Getting the plotting data
ploting_data_passeriformes <- list(
                     "median_elaborations" = apply(get.disparity(get.subsets(results$suborder_results_phylo$position, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
                     "median_innovations"  = apply(get.disparity(get.subsets(results$suborder_results_phylo$distance, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median),
                     "distances_to_centre" = get.disparity(dispRity(get.subsets(suborder_shapespace, subsets = "phylogeny"), metric = dispRity::centroids, centroid = 0))[[1]],
                     "consensus_tree"      = tree) 

## Get the data at the macro level (with no missing data)
get.macro.projection <- function(data, what) {
    if(what == "centroids") {
        ## Measuring the centroid distances
        data <- dispRity(data, metric = dispRity::centroids)
    } else {
        data <- data[[what]]
    }
    ## Getting all the species
    all_species <- apply(get.disparity(get.subsets(data, subsets = "phylogeny"), concatenate = FALSE)[[1]], 1, median)
    names(all_species) <- rownames(get.subsets(data, subsets = "phylogeny")$matrix[[1]])
    ## Getting the species that have a projection on their own group
    macro_level <- apply(do.call(rbind, get.disparity(get.subsets(data, subsets = names(size.subsets(data))[-n.subsets(data)]), concatenate = FALSE)), 1, median)
    names(macro_level) <- unlist(lapply(lapply(get.subsets(data, subsets = names(size.subsets(data))[-n.subsets(data)])$subsets, `[[`, "elements"), function(X, matrix) rownames(matrix)[X], matrix = data$matrix[[1]]))
    ## Replacing the species that have their own group
    all_species[match(names(macro_level), names(all_species))] <- macro_level
    return(all_species)
}

## Getting the plotting data super_order_level
ploting_data_passeriformes_suborder <- list(
                     "median_elaborations" = get.macro.projection(results$suborder_results_within, what = "position"),
                     "median_innovations"  = get.macro.projection(results$suborder_results_within, what = "distance"),
                     "distances_to_centre" = get.macro.projection(suborder_shapespace, what = "centroids"),
                     "consensus_tree"      = tree) 

## Getting the plotting data super_order_level
ploting_data_passeriformes_family <- list(
                     "median_elaborations" = get.macro.projection(results$family_results_within, what = "position"),
                     "median_innovations"  = get.macro.projection(results$family_results_within, what = "distance"),
                     "distances_to_centre" = get.macro.projection(family_shapespace, what = "centroids"),
                     "consensus_tree"      = tree) 

## Save all that as a list
save(ploting_data_passeriformes, file = "../Data/Processed/ploting_data_passeriformes.rda")
## The projections within the suborders
save(ploting_data_passeriformes_suborder, file = "../Data/Processed/ploting_data_passeriformes_suborder.rda")
## The projections within the families
save(ploting_data_passeriformes_family, file = "../Data/Processed/ploting_data_passeriformes_family.rda")

## Group info
suborders <- lapply(results$suborder_results_phylo$position$subsets, function(x, data) list(species = rownames(data)[c(x$elements)]), data = results$suborder_results_phylo$position$matrix[[1]])
families <- lapply(results$family_results_phylo$position$subsets, function(x, data) list(species = rownames(data)[c(x$elements)]), data = results$family_results_phylo$position$matrix[[1]])
## Adding the colours
load("../Data/Processed/tip_colours_suborders.rda")
load("../Data/Processed/tip_colours_families.rda")
add.col <- function(one_order, one_col) {
    one_order$col <- one_col
    return(one_order)
}
suborders <- mapply(add.col, one_order = suborders[1:8][match(names(suborders[1:8]), names(tip_colours_suborders))], one_col = as.list(tip_colours_suborders), SIMPLIFY = FALSE)
families <- mapply(add.col, one_order = families[1:27][match(names(families[1:27]), names(tip_colours_families))], one_col = as.list(tip_colours_families), SIMPLIFY = FALSE)

## Big list:
clade_info_passeriformes <- list(suborders = suborders,
                                 families = families)
## Save all that data
save(clade_info_passeriformes, file = "../Data/Processed/clade_info_passeriformes.rda")
```




































