---
title: "Bird beak's elaboration and innovation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

This section contains the code for reproducing the analyses. To get the details, refer to [`dispRity` vignette](link_to_projection_vignette).
This script take around 6 minutes to compile.

```{r, eval = TRUE, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, print = FALSE}
library(dispRity)
source("../Functions/ploting.utilities.R")
```

# Measuring ellipse's statistics

For each group, we can calculate:

 * It's (dis)alignment relative to the the phylogeny
 * It's distance from the centroid of the phylogeny
 * It's angles sd
 * It's orientation (angle) relative to the whole phylogeny
 * Whether that orientation is significant


# Loading the data

```{r, eval = TRUE, echo = FALSE}
## Load the data for all birds
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
## Select the data
shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

## Load the covar
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]

## Get the levels
phylo_lvl <- which(MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order) == "animal")
super_lvl <- grep("level1_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
order_lvl <- grep("level2_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))

## Create the subset for all birds
all_birds_shapespace <- MCMCglmm.subsets(n = 500, # Just run it on 500 matrices for now
      data          = shapespace,
      posteriors    = shapespace_allbirds_lvl_superorder_order,
      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(super_lvl, order_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)


# Load the data for the passeriformes
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
## Select the data
shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

## Load the covar
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]

## Get the levels
phylo_lvl <- which(MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order) == "animal")
super_lvl <- grep("level1_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
order_lvl <- grep("level2_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))

## Create the subset for all birds
all_birds_shapespace <- MCMCglmm.subsets(n = 500, # Just run it on 500 matrices for now
      data          = shapespace,
      posteriors    = shapespace_allbirds_lvl_superorder_order,
      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(super_lvl, order_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level1)[-1], levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)




```





### Groups compared to the phylogeny

```{r, eval = FALSE, echo = FALSE}
source("../Functions/stats.utilities.R")
## General measurement
super_orders_stats <- ellipse.stats(superorder_shapespace)
orders_stats <- ellipse.stats(order_shapespace)
suborder_stats <- ellipse.stats(suborder_shapespace)
family_stats <- ellipse.stats(family_shapespace)

ellipse_stats <- list(super_orders_stats = super_orders_stats, 
                      orders_stats = orders_stats,
                      suborder_stats = suborder_stats,
                      family_stats = family_stats)
save(ellipse_stats, file = "../Data/Processed/ellipse_stats.rda")
```






```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/ellipse_stats.rda")
## Ploting fun
ploting.stats <- function(results, col) {

    col <- col[match(names(size.subsets(results[[1]])), names(col))]

    layout_matrix <- layout(matrix(1:3, nrow = 1), widths = c(4.2, 3, 3), heights = n.subsets(results[[1]])/2)
    par(mar = c(5, 10, 2, 1) + 0.1)
    plot(results[[1]], horizontal = TRUE, las = 1, xlab = "angle", ylab = "", col = col)
    par(mar = c(5, 1, 2, 1) + 0.1)
    plot(results[[2]], horizontal = TRUE, las = 1, xlab = "disalignment", ylab = "", yaxt = "n", col = col)
    par(mar = c(5, 1, 2, 1) + 0.1)
    plot(results[[3]], horizontal = TRUE, las = 1, xlab = "distance between centroids", ylab = "", yaxt = "n", col = col)
}

## Ploting fun
table.stats <- function(results, class) {
    ## Table results
    table <- cbind(summary(results[[1]])[, -c(1:3)], summary(results[[2]])[, -c(1:3)], summary(results[[3]])[, -c(1:3)])
    rownames(table) <- gsub(":phy", "", as.vector(summary(results[[1]])[,1]))
    colnames(table)[c(1,6,11)] <- c("angle", "disalignment", "distance")
    if(!missing(class)) {
        table <- cbind(class = class, table)
    }
    return(table)
}

all_results <- rbind(table.stats(ellipse_stats$super_orders_stats, class = "superorder"),
                     table.stats(ellipse_stats$orders_stats, class = "orders"),
                     table.stats(ellipse_stats$suborder_stats, class = "suborder"),
                     table.stats(ellipse_stats$family_stats, class = "family"))
knitr::kable(all_results)
```










```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/tip_colours_super_orders.rda")
ploting.stats(ellipse_stats$super_orders_stats, col = tip_colours_super_orders)
```

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/tip_colours_orders.rda")
ploting.stats(ellipse_stats$orders_stats, col = tip_colours_orders)
```

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/tip_colours_passeriformes_suborder.rda")
ploting.stats(ellipse_stats$suborder_stats, col = tip_colours_passeriformes_suborder)
```

```{r, eval = TRUE, echo = FALSE}
load("../Data/Processed/tip_colours_passeriformes_family.rda")
ploting.stats(ellipse_stats$family_stats, col = tip_colours_passeriformes_family)
```



