---
title: "Bird beak's elaboration and innovation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

This script reproduces a figure not present or used in the manuscript or the supplementary materials.

```{r, eval = TRUE, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, print = FALSE}
library(dispRity)
source("../Functions/ploting.utilities.R")
source("../Functions/correlations.utilities.R")
```

# All birds analysis

## Loading the data

```{r, eval = TRUE, echo = TRUE}
## Load the data
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
## Select the data
shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

## Load the covar
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]

## Get the levels
phylo_lvl <- which(MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order) == "animal")
super_lvl <- grep("level1_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
order_lvl <- grep("level2_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))

## Creating the subsets
superorder_shapespace <- MCMCglmm.subsets(, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_allbirds_lvl_superorder_order,
      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(super_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""

order_shapespace <- MCMCglmm.subsets(, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_allbirds_lvl_superorder_order,
      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(order_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

## Loading the projection/rejections results
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
results <- shapespace_allbirds_lvl_superorder_order_results_list
rm(shapespace_allbirds_lvl_superorder_order_results_list)
```

Extract the median elaboration and innovation from the results for the mega level (projection on the phylogeny) and the macro level (projection on each group).

```{r, eval = TRUE, echo = TRUE}
## Calculate the correlations for the projections onto the phylo axis
super_phylo_cor <- get.correlations.plot(results$super_results_phylo)

## Calculate the correlations for the projections onto the group' axes
super_group_cor <- get.correlations.plot(results$super_results_within, exclude.phylo = TRUE)

## Check for order_results_within
order_correlations <- get.correlations.plot(results$order_results_within, exclude.phylo = TRUE)
```

## Get the correlation fable


```{r, echo = TRUE, eval = TRUE}
## Get all the correlations for all the clades
cor_mega_super_orders <- get.correlations.scores(results$super_results_phylo)
cor_macro_super_orders <- get.correlations.scores(results$super_results_within)
cor_macro_orders <- get.correlations.scores(results$order_results_within)

## Get the colours
load("../Data/Processed/tip_colours_super_orders.rda")
super_orders_col <- c(tip_colours_super_orders, "phylogeny" = "grey")

## Plot the correlations
pdf("../Manuscript/Figures/correlations_fable_super_orders.pdf")
par(mfrow = c(1,2))
par(mar = c(5, 8, 4, 0) + 0.1, bty = "n") # c(bottom, left, top, right)
plot.cor.scores(cor_mega_super_orders, col = super_orders_col,
                y.axis = TRUE, plot.phylo = TRUE,
                main = "Species projected onto\nthe whole phylogeny major axes") 
par(mar = c(5, 2, 4, 6) + 0.1, bty = "n") # c(bottom, left, top, right)
plot.cor.scores(cor_macro_super_orders, col = super_orders_col,
                y.axis = FALSE, plot.phylo = FALSE,
                main = "Species projected onto\ntheir superorders' major axes")
dev.off()


## Correlation fable
pdf(width = 5, height = 9, file = "../Manuscript/Figures/correlations_fable_orders.pdf")
load("../Data/Processed/tip_colours_orders.rda")
orders_col <- c(tip_colours_orders, "phylogeny" = "grey")
orders_col[] <- "black"
par(mar = c(5, 8, 4, 0) + 0.1, bty = "n") # c(bottom, left, top, right)
plot.cor.scores(cor_macro_orders, col = orders_col,
                y.axis = TRUE, plot.phylo = FALSE,
                show.phylo = FALSE, xlim = c(-0.5, 1),
                main = "Species projected onto\ntheir orders' major axes") 
dev.off()
```




## Ploting the figure

The following snippets plots the entire figure minus the panel IDs:
 * Panel A: the overal density figure
 * Panel B: the correlation plots for the projection on the mega level
 * Panel C: the correlation plots for the projection on the macro level

Density plot of the correlation of the median elaboration and median exploration for all species.

```{r, echo = TRUE, eval = TRUE}
## Set the pdf parameters
pdf("../Manuscript/Figures/correlations.pdf", height = 12, width = 8)

## Set up the plotting layout
layout_matrix <- cbind(matrix(1,2,2), matrix(0,2,2))
layout_matrix <- rbind(rbind(layout_matrix, matrix(1:8+1, 2, 4, byrow = TRUE)), matrix(9:16+1, 2, 4,  byrow = TRUE))
plot_layout <- layout(layout_matrix)
# layout.show(plot_layout)

## Setting the colour pallette for the subsequent plots
colour.palette <- gg.color.hue

## --------
## PANEL A
## --------

## Plotting the densities (with the quadrants) in the first panel
par(mar = c(5, 4, 4, 2) + 0.1)
plot.densities(super_phylo_cor$phylogeny, with.quadrant = TRUE)

## --------
## PANEL B
## --------

## Getting the subsets for the colouring
col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))

## Set the vector of legend positions
legends_pos <- c("topleft", "topright", "topright", "topright",
                 "topleft", "topright", "topleft", "topright")
## Create the vector for counting the legend numbers
legend_counter <- 1

## Plotting each individual correlation plots
for(i in 1:8) {
    par(mar = c(2,2.5,2,1.5))
    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = TRUE)
}

## --------
## PANEL C
## --------

## Set the vector of legend positions
legends_pos <- c("bottomright", "topright", "topleft", "topleft",
                 "topleft", "topleft", "topright", "bottomright")
## Resetting the legend_counter
legend_counter <- 1

## Plotting each individual correlation plots
for(i in 1:8) {
    par(mar = c(2,2.5,2,1.5))
    plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
}
dev.off()


## Extra figure
pdf("../Manuscript/Figures/correlations_2.pdf", height = 20/3*1.3, width = 8*1.3)
par(mfrow = c(5,6))
for(i in 1:26) {
    par(mar = c(2,4,2.5,0)+0.1)
    plot.correlations(order_correlations[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
}
dev.off()
```

Extract the median elaboration and innovation from the results for the macro level (projection on the passeriformes) and the macro level-1 (projection on each group).

# Passeriformes analysis

## Loading the data

```{r, eval = TRUE, echo = TRUE}
## Load the data
load(file = "../Data/Processed/shapespace_passeriformes_two_levels.rda")
## Select the data
shapespace <- shapespace_passeriformes_two_levels[[1]]$space
tree <- shapespace_passeriformes_two_levels[[1]]$consensus_tree
## Clean garbage
rm(shapespace_passeriformes_two_levels)

## Load the covar
load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
shapespace_passeriformes <- shapespace_passeriformes[[1]]

## Get the levels
phylo_lvl <- which(MCMCglmm.levels(shapespace_passeriformes) == "animal")
subor_lvl <- grep("level1_", MCMCglmm.levels(shapespace_passeriformes))
famil_lvl <- grep("level2_", MCMCglmm.levels(shapespace_passeriformes))

## Creating the subsets
suborder_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_passeriformes,
      group         = MCMCglmm.levels(shapespace_passeriformes)[c(subor_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""

family_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_passeriformes,
      group         = MCMCglmm.levels(shapespace_passeriformes)[c(famil_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_passeriformes)

## Loading the results
load(file = "../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")
results <- shapespace_passeriformes_suborder_family_results_list
rm(shapespace_passeriformes_suborder_family_results_list)
```


```{r, eval = TRUE, echo = FALSE}
## Calculate the correlations for the projections onto the phylo axis
suborder_phylo_cor <- get.correlations.plot(results$suborder_results_phylo)

## Calculate the correlations for the projections onto the group' axes
suborder_group_cor <- get.correlations.plot(results$suborder_results_within, exclude.phylo = TRUE)
```

## Ploting the figure

The following snippets plots the entire figure minus the panel IDs:
 * Panel A: the overal density figure
 * Panel B: the correlation plots for the projection on the mega level
 * Panel C: the correlation plots for the projection on the macro level


Density plot of the correlation of the median elaboration and median exploration for all species.

```{r, echo = TRUE, eval = TRUE}
## Set the pdf parameters
pdf("../Manuscript/Figures/correlations_passeriformes.pdf", height = 12, width = 6)

## Set up the plotting layout
layout_matrix <- cbind(matrix(1,2,2), matrix(0,2,1))
layout_matrix <- rbind(rbind(layout_matrix, matrix(1:6+1, 2, 3, byrow = TRUE)), matrix(7:12+1, 2, 3,  byrow = TRUE))
plot_layout <- layout(layout_matrix)
# layout.show(plot_layout)

## Setting the colour pallette for the subsequent plots
colour.palette <- gg.color.hue

## --------
## PANEL A
## --------

## Plotting the densities (with the quadrants) in the first panel
par(mar = c(5, 4, 4, 2) + 0.1)
plot.densities(suborder_phylo_cor$phylogeny, with.quadrant = TRUE)

## --------
## PANEL B
## --------

## Getting the subsets for the colouring
col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))

## Set the vector of legend positions
legends_pos <- c("topright", "topright", "topright", "topright", "topright", "topright")
## Create the vector for counting the legend numbers
legend_counter <- 1

## Plotting each individual correlation plots
for(i in 1:6) {
    par(mar = c(2,2.5,2,1.5))
    plot.correlations(suborder_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = TRUE)
}

## --------
## PANEL C
## --------

## Set the vector of legend positions
legends_pos <- c("topright", "topright", "topright", "topright", "topright", "topright")

## Resetting the legend_counter
legend_counter <- 1

## Plotting each individual correlation plots
for(i in 1:6) {
    par(mar = c(2,2.5,2,1.5))
    plot.correlations(suborder_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
}
dev.off()
```

Legend numbers are:
1 - others (groups with < 15 species)
2 - Corvoidea
3 - Malaconotoidea
4 - Orioloidea
5 - Meliphagoidea
6 - Bombycilloidea
7 - Muscicapoidea
8 - Emberizoidea
9 - Motacillidae
10 - Nectariniidae 
11 - Passeridae
12 - PloceidaeEstrildidae
13 - Eurylaimides
14 - Furnariida
15 - Tyrannida 
16 - Aegithaloidea
17 - Cisticolidae
18 - Fringillidae
19 - Hirundinidae
20 - Locustelloidea
21 - Paridae
22 - Pycnonotidae
23 - Sylvioidea



















































