---
title: "Bird beak's elaboration and innovation"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    fig_width: 8
    fig_height: 8
---

This markdown contains the code to reproduce the figures containing the elaboration and innovation per species for the all birds and the passeriformes analyses.
For the details on how to get the data objects called here, refer to `03-elaboration_innovation_analyses.Rmd`.


```{r, eval = TRUE, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE, print = FALSE}
library(dispRity)
source("../Functions/ploting.utilities.R")
source("../Functions/correlations.utilities.R")
```

# All birds analysis

## Loading the data

```{r, eval = TRUE, echo = TRUE}
## Load the data
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.rda")
## Select the data
shapespace <- shapespace_allbirds_lvl_superorder_order[[1]]$space
tree <- shapespace_allbirds_lvl_superorder_order[[1]]$consensus_tree
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

## Load the covar
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order.MCMCglmm")
shapespace_allbirds_lvl_superorder_order <- shapespace_allbirds_lvl_superorder_order[[1]]

## Get the levels
phylo_lvl <- which(MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order) == "animal")
super_lvl <- grep("level1_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))
order_lvl <- grep("level2_", MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order))

## Creating the subsets
superorder_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_allbirds_lvl_superorder_order,
      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(super_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""

order_shapespace <- MCMCglmm.subsets(n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_allbirds_lvl_superorder_order,
      group         = MCMCglmm.levels(shapespace_allbirds_lvl_superorder_order)[c(order_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_allbirds_lvl_superorder_order)

## Loading the projection/rejections results
load(file = "../Data/Processed/shapespace_allbirds_lvl_superorder_order_results_list.rda")
results <- shapespace_allbirds_lvl_superorder_order_results_list
rm(shapespace_allbirds_lvl_superorder_order_results_list)
```

Extract the median elaboration and innovation from the results for the mega level (projection on the phylogeny) and the macro level (projection on each group).

```{r, eval = TRUE, echo = TRUE}
## Calculate the correlations for the projections onto the phylo axis
super_phylo_cor <- get.correlations(results$super_results_phylo)

## Calculate the correlations for the projections onto the group' axes
super_group_cor <- get.correlations(results$super_results_within, exclude.phylo = TRUE)
```

## Ploting the figure

The following snippets plots the entire figure minus the panel IDs:
 * Panel A: the overal density figure
 * Panel B: the correlation plots for the projection on the mega level
 * Panel C: the correlation plots for the projection on the macro level


Density plot of the correlation of the median elaboration and median exploration for all species.

```{r, echo = TRUE, eval = TRUE}
## Set the pdf parameters
pdf("../Manuscript/Figures/Figure2_correlations.pdf", height = 12, width = 8)

## Set up the plotting layout
layout_matrix <- cbind(matrix(1,2,2), matrix(0,2,2))
layout_matrix <- rbind(rbind(layout_matrix, matrix(1:8+1, 2, 4, byrow = TRUE)), matrix(9:16+1, 2, 4,  byrow = TRUE))
plot_layout <- layout(layout_matrix)
# layout.show(plot_layout)

## Setting the colour pallette for the subsequent plots
colour.palette <- gg.color.hue

## --------
## PANEL A
## --------

## Plotting the densities (with the quadrants) in the first panel
par(mar = c(5, 4, 4, 2) + 0.1)
plot.densities(super_phylo_cor$phylogeny, with.quadrant = TRUE)

## --------
## PANEL B
## --------

## Getting the subsets for the colouring
col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))

## Set the vector of legend positions
legends_pos <- c("topleft", "topright", "topright", "topright",
                 "topleft", "topright", "topleft", "topright")
## Create the vector for counting the legend numbers
legend_counter <- 1

## Plotting each individual correlation plots
for(i in 1:8) {
    par(mar = c(2,2.5,2,1.5))
    plot.correlations(super_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = TRUE)
}

## --------
## PANEL C
## --------

## Set the vector of legend positions
legends_pos <- c("bottomright", "topright", "topleft", "topleft",
                 "topleft", "topleft", "topright", "bottomright")
## Resetting the legend_counter
legend_counter <- 1

## Plotting each individual correlation plots
for(i in 1:8) {
    par(mar = c(2,2.5,2,1.5))
    plot.correlations(super_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
}
dev.off()
```






Extract the median elaboration and innovation from the results for the macro level (projection on the passeriformes) and the macro level-1 (projection on each group).



# Passeriformes analysis

## Loading the data

```{r, eval = TRUE, echo = TRUE}
## Load the data
load(file = "../Data/Processed/shapespace_passeriformes_two_levels.rda")
## Select the data
shapespace <- shapespace_passeriformes_two_levels[[1]]$space
tree <- shapespace_passeriformes_two_levels[[1]]$consensus_tree
## Clean garbage
rm(shapespace_passeriformes_two_levels)

## Load the covar
load(file = "../Data/Processed/shapespace_passeriformes.MCMCglmm")
shapespace_passeriformes <- shapespace_passeriformes[[1]]

## Get the levels
phylo_lvl <- which(MCMCglmm.levels(shapespace_passeriformes) == "animal")
subor_lvl <- grep("level1_", MCMCglmm.levels(shapespace_passeriformes))
famil_lvl <- grep("level2_", MCMCglmm.levels(shapespace_passeriformes))

## Creating the subsets
suborder_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_passeriformes,
      group         = MCMCglmm.levels(shapespace_passeriformes)[c(subor_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level1)[-1], "phylogeny")) #[-1] is because the first level is ""

family_shapespace <- MCMCglmm.subsets(#n = 1000, # Just run it on 1k matrices for now
      data          = shapespace,
      posteriors    = shapespace_passeriformes,
      group         = MCMCglmm.levels(shapespace_passeriformes)[c(famil_lvl, phylo_lvl)],
      rename.groups = c(levels(shapespace$level2)[-1], "phylogeny")) #[-1] is because the first level is ""
## Clean garbage
rm(shapespace_passeriformes)

## Loading the results
load(file = "../Data/Processed/shapespace_passeriformes_suborder_family_results_list.rda")
results <- shapespace_passeriformes_suborder_family_results_list
rm(shapespace_passeriformes_suborder_family_results_list)
```


```{r, eval = TRUE, echo = FALSE}
## Calculate the correlations for the projections onto the phylo axis
suborder_phylo_cor <- get.correlations(results$suborder_results_phylo)

## Calculate the correlations for the projections onto the group' axes
suborder_group_cor <- get.correlations(results$suborder_results_within, exclude.phylo = TRUE)
```

## Ploting the figure

The following snippets plots the entire figure minus the panel IDs:
 * Panel A: the overal density figure
 * Panel B: the correlation plots for the projection on the mega level
 * Panel C: the correlation plots for the projection on the macro level


Density plot of the correlation of the median elaboration and median exploration for all species.

```{r, echo = TRUE, eval = TRUE}
## Set the pdf parameters
pdf("../Manuscript/Figures/Figure2_correlations_passeriformes.pdf", height = 12, width = 6)

## Set up the plotting layout
layout_matrix <- cbind(matrix(1,2,2), matrix(0,2,1))
layout_matrix <- rbind(rbind(layout_matrix, matrix(1:6+1, 2, 3, byrow = TRUE)), matrix(7:12+1, 2, 3,  byrow = TRUE))
plot_layout <- layout(layout_matrix)
# layout.show(plot_layout)

## Setting the colour pallette for the subsequent plots
colour.palette <- gg.color.hue

## --------
## PANEL A
## --------

## Plotting the densities (with the quadrants) in the first panel
par(mar = c(5, 4, 4, 2) + 0.1)
plot.densities(suborder_phylo_cor$phylogeny, with.quadrant = TRUE)

## --------
## PANEL B
## --------

## Getting the subsets for the colouring
col_subs <- c(split(shapespace$level2, f = shapespace$level1)[-1], "phylogeny" = list(shapespace$level1))

## Set the vector of legend positions
legends_pos <- c("topright", "topright", "topright", "topright", "topright", "topright")
## Create the vector for counting the legend numbers
legend_counter <- 1

## Plotting each individual correlation plots
for(i in 1:6) {
    par(mar = c(2,2.5,2,1.5))
    plot.correlations(suborder_phylo_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, contour = NULL, with.quadrant = TRUE)
}

## --------
## PANEL C
## --------

## Set the vector of legend positions
legends_pos <- c("topright", "topright", "topright", "topright", "topright", "topright")

## Resetting the legend_counter
legend_counter <- 1

## Plotting each individual correlation plots
for(i in 1:6) {
    par(mar = c(2,2.5,2,1.5))
    plot.correlations(suborder_group_cor[[i]], col.sub = col_subs[[i]], legend.pos = legends_pos[i], ID = i, pt.cex = 0.5, legend.cex = 2/3, legend_counter = legend_counter, with.quadrant = FALSE, contour = NULL)
}
dev.off()
```

Legend numbers are:
1 - others (groups with < 15 species)
2 - Corvoidea
3 - Malaconotoidea
4 - Orioloidea
5 - Meliphagoidea
6 - Bombycilloidea
7 - Muscicapoidea
8 - Emberizoidea
9 - Motacillidae
10 - Nectariniidae 
11 - Passeridae
12 - PloceidaeEstrildidae
13 - Eurylaimides
14 - Furnariida
15 - Tyrannida 
16 - Aegithaloidea
17 - Cisticolidae
18 - Fringillidae
19 - Hirundinidae
20 - Locustelloidea
21 - Paridae
22 - Pycnonotidae
23 - Sylvioidea



















































