---
title: "Updates a Procrustes and a PCA"
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 6
    fig_height: 6
---

In this script we will be looking at how to update both a Procrustes superimposition (GPA) and a PCA by projecting new specimens in it.
The goal is to use the GPA and PCA from a previous dataset (published with Guillerme et al 2023 - Science Advances) and to project new specimens onto it.

# The pipeline (testing)

Here is an example on how to do it using the `Morpho` package for the base GPA and PCA:

```{r}
library(Morpho)

## Get some 3D data from Morpho
data(boneData)
original_landmarks <- boneLM

## Run the GPA for the original data
proc_orig <- Morpho::procSym(original_landmarks)

## Run the PCA for the original data
proc_matrix_orig <- landvR::array.to(proc_orig$rotated, to = "matrix")
pca_orig <- prcomp(proc_matrix_orig)

## Plot the results
plot(pca_orig$x[,c(1,2)])
```

And now let's generate some extra data out of the pipeline.
Here we're just creating a mean, median and minimum landmark set using the mean and median landmark coordinates values as well as one of a specific specimen (42).

```{r}
## Create a new specimens: the average, the median and the minimum, of all bones)
new_spec <- array(data = c(apply(original_landmarks, c(1,2), mean),
                           apply(original_landmarks, c(1,2), median),
                           original_landmarks[,,42]), dim = c(10, 3, 3))
dimnames(new_spec)[[3]] <- paste0("new_", c("mean", "median", "spec42"))

## Here is what the landmark positions look like in 3D
op <- par(mfrow = c(2,2))
plot(original_landmarks[, c(1,2), 1], main = "First specimen\n(reference from real dataset)")
plot(new_spec[, c(1,2), 1], main = "New specimen 1\n(average)")
plot(new_spec[, c(1,2), 2], main = "New specimen 2\n(median)")
plot(new_spec[, c(1,2), 3], main = "New specimen 3\n(spec 42)")
par(op)
```

We can then add these specimens to the GPA and PCA.

```{r}
## Updating the gpa and pca
updated_pca <- update.gpa.pca(proc_orig, new_spec)

## Plot all the results
plot(proc_orig$PCscores[,c(1,2)])
points(updated_pca[-c(81:83), c(1,2)], pch = 19, cex = 0.5)
points(updated_pca[c(81:83), c(1,2)], pch = 19, col = "red")
legend("topright", pch = c(21, 19, 19), col = c("black", "black", "red"), pt.cex = c(1, 0.5, 1), legend = c("original PCA", "updated PCA", "new specimen"))
```

Seems to work quiet well!
All the positions from previous specimens are respected.
And the two new median and average specimens plot in the centre of the traitspace (as expected) with the weird one plotting on some edge of the traitspace.

The function that does it automatically is now implemented in `landvR`.

# Mapping the chaffinches on the whole traitspace

```{r}
## Loading the entire bird GPA
library(Morpho)
whole_gpa <- readRDS("../Data/Raw/Beak_data/2020_08_07_MMB_MORPHO_SHAPESPACE_FULL.rds")

## Reading in the chaffinches
land_pairs <- read.table("../Data/Raw/Beak_data/landpairs.txt")

## Read all the chaffinches
path <- "../Data/Raw/Beak_data/Chaffinch_data/"
individuals <- list.files(path)
all_individuals <- list()
while(length(individuals) != 0) {
    file <- paste0(path, individuals[1], "/", list.files(paste0(path, individuals[1]))) 
    if(grepl(".nts", file)) { 
        ## Does contain a file
        data <- readland.nts(paste0(path, individuals[1], "/", list.files(paste0(path, individuals[1]))))
        name <- paste0(strsplit(individuals[1], split = "_")[[1]][1:4], collapse = "_")
        all_individuals[[length(all_individuals) + 1]] <- data[, , 1]
        names(all_individuals)[length(all_individuals)] <- name
    } #else {
        # print(paste0(file, ": empty"))
    # }
    individuals <- individuals[-1]
}

## The following are missing:
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_africana_14_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_africana_19_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_coelebs_1_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_gengleri_1_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_gengleri_21_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_gengleri_5_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_moreletti_11_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_coelebs_palmae_13_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_teydea_polatzeki_1_intraspecific/: empty"
# [1] "../Data/Raw/Beak_data/Chaffinch_data/Fringilla_teydea_teydea_10_intraspecific/: empty"

## Converting into an array
individuals_array <- simplify2array(all_individuals)

## Project them! (*-1 is to flip it around)
updated_pca <- update.gpa.pca(whole_gpa, individuals_array, scale.pca = FALSE)

## BUG IN THE NUMBER OF DIMENSIONS!
dim(updated_pca)[1] == dim(whole_gpa$PCscores)[1] + length(all_individuals)
dim(updated_pca)[2] == dim(whole_gpa$PCscores)[2]

## SOME SMALL SHIFT!
plot(whole_gpa$PCscores[,c(1,2)], col = "grey")
points(updated_pca[-c((dim(whole_gpa$PCscores)[1]+1): dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "grey")
## Plot the chaffinches to see where they lay:
fringillas <- which(grepl("Fringilla", rownames(whole_gpa$PCscores)))
points(whole_gpa$PCscores[fringillas,c(1,2)], pch = 21, col = "blue")
points(updated_pca[c((dim(whole_gpa$PCscores)[1]+1):dim(updated_pca)[1]), c(1,2)], pch = 19, cex = 0.5, col = "orange")
legend("bottomright", pch = c(21, 19, 21, 19), col = c("grey", "grey", "blue", "orange"), pt.cex = c(1, 0.5, 1, 0.5), legend = c("original PCA", "updated PCA", "Fringilla (original)", "New specimens"))
```
